# Length data {#length}

## Preamble
___

Load libraries and data:

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(lubridate)
library(tidyices)
library(ggridges)
```

```{r}
hh <- read_rds("data/ns-ibts_hh.rds")
hl <- read_rds("data/ns-ibts_hl.rds")
```

## Catch per haul

### Abundance

Since in the tidying process all length data were standardized to 60 minutes haul time it is easy to calculate the standardized abundance per haul. Lets do this for some species of choice:

```{r}
Latin <- "Pleuronectes platessa"
by.haul.positive <- 
  hl %>% 
  select(-sex) %>% 
  filter(latin == Latin) %>% 
  drop_na() %>% 
  group_by(id) %>% 
  summarise(n = sum(n))
```

In the code above we have basically collapsed the abundance by length to abundance by haul. The number of records in the by.haul dataframe are `r nrow(by.haul.positive)` while the number of hauls in the haul dataframe are `r nrow(hh)`. The difference are hauls where no `r Latin` was caught.

In order to include the stations with zero catch we:

* Take the haul data and do a left join with the length data which have some catch recorded. A demonstration of the join-function family is provided using some simple data in [On joins](#joins).
* This results in the zero stations having NA in the n variable. Hence the mutate in combination with replace_na comes to the rescue, setting a value of zero as count for stations where no fish was caught otherwise retain the count value.

```{r}
by.haul <-
  hh %>% 
  select(id, year, quarter, haulval) %>% 
  left_join(by.haul.positive) %>%
  mutate(n = replace_na(n, 0))
```

A quick count:
```{r}
table(by.haul$haulval, by.haul$n == 0)
```

reveals that we have `r by.haul %>% filter(haulval == "V", n > 0) %>% nrow()` valid hauls with `r Latin` and `r by.haul %>% filter(haulval == "V", n == 0) %>% nrow()` valid hauls where no `r Latin` was caught.

Since we are here we may as well look at the temporal trend in the median abundance using only the valid hauls:

```{r}
by.haul %>% 
  filter(haulval == "V") %>% 
  group_by(year, quarter) %>% 
  summarise(m = mean(n)) %>% 
  ungroup() %>% 
  mutate(year = year + 0.25 * quarter - 0.125) %>% 
  ggplot(aes(year, m, colour = factor(quarter), group = quarter)) +
  geom_point() +
  geom_line() +
  labs(x = NULL, y = NULL,
       colour = "Quarter",
       title = paste("Mean number of", Latin, "per 1 hour haul"))
```

Using the mean may not be appropriate here, given the general distribution of abundance by haul one observes in groundfish surveys:

```{r}
by.haul %>% 
  filter(haulval == "V") %>% 
  # put the values above the 99%th percentile to 99%th percentile
  mutate(n = ifelse(n > quantile(n, 0.99), quantile(n, 0.99), n)) %>% 
  ggplot(aes(n)) +
  geom_histogram() +
  labs(x = "Abundance per haul",
       y = "Number of stations")
```

Here we have most hauls with relatively small or zero catches with occational hauls with very large catches resulting in a highly negatively scewed distribution. The ggplot2 package has a very nice summary statistics (bootstrap mean and standard error) that comes to the rescue:

```{r}
by.haul %>% 
  filter(haulval == "V") %>% 
  mutate(year = year + 0.25 * quarter - 0.125) %>% 
  ggplot(aes(year, n, colour = factor(quarter))) +
  stat_summary(fun.data = "mean_cl_boot", size = 0.1) +
  expand_limits(y = 0) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = NULL,
       colour = "Quarter",
       title = paste("Mean number of", Latin, "per 1 hour haul"),
       subtitle = "Bootstrap mean and one standard error")
```

### Weigth (biomass)

If interested in biomass trend the code can easily be ammended, using some isometric length-weight parameters as a rough guess (we will obtain the proper one in XXX). We need to start from scratch and mutate length to weight before summarising by weight for each haul:

```{r}
by.haul.positive <- 
  hl %>% 
  select(-sex) %>% 
  filter(latin == Latin) %>% 
  drop_na() %>% 
  # additional code
  mutate(wt = n * 0.00001 * length^3) %>% 
  group_by(id) %>% 
  summarise(n = sum(n),
            # additional code
            wt = sum(wt))

by.haul <-
  hh %>% 
  select(id, year, quarter, haulval) %>% 
  left_join(by.haul.positive) %>% 
  mutate(n  = replace_na(n,  0),
         # additional code
         wt = replace_na(wt, 0))
by.haul %>% 
  filter(haulval == "V") %>% 
  mutate(year = year + 0.25 * quarter - 0.125) %>% 
  # Note: Here only replaced variable "n" with "wt":
  ggplot(aes(year, wt, colour = factor(quarter))) +
  stat_summary(fun.data = "mean_cl_boot", size = 0.1) +
  expand_limits(y = 0) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = NULL,
       colour = "Quarter",
       title = paste("Mean catch [kg] of", Latin, "per 1 hour haul"),
       subtitle = "Bootstrap mean and one standard error")
```

### More species

With some yet another minor tweaking we can acutally do this for all species. Here we will though limit ourselves to some "common" species:

```{r}
Latin <- c("Clupea harengus", "Sprattus sprattus", "Gadus morhua",
           "Melanogrammus aeglefinus", "Merlangius merlangus", "Pollachius virens",
           "Trisopterus esmarkii", "Scomber scombrus", "Pleuronectes platessa")

by.haul.positive <- 
  hl %>% 
  select(-sex) %>% 
  filter(latin %in% Latin) %>% 
  drop_na() %>% 
  # additional code
  mutate(wt = n * 0.00001 * length^3) %>% 
  # additional code: added latin to the grouping
  group_by(id, latin) %>% 
  summarise(n = sum(n),
            # additional code
            wt = sum(wt))

by.haul <-
  hh %>% 
  select(id, year, quarter, haulval) %>% 
  left_join(by.haul.positive) %>% 
  mutate(n  = replace_na(n,  0),
         # additional code
         wt = replace_na(wt, 0))
by.haul %>% 
  filter(haulval == "V") %>% 
  mutate(year = year + 0.25 * quarter - 0.125) %>% 
  # NOTE: Check why this step is
  drop_na() %>% 
  # Note: Here only replaced variable "n" with "wt":
  ggplot(aes(year, wt, colour = factor(quarter))) +
  stat_summary(fun.data = "mean_cl_boot", size = 0.1) +
  expand_limits(y = 0) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = NULL,
       colour = "Quarter",
       title = "Mean catch [kg] per 1 hour haul",
       subtitle = "Bootstrap mean and one standard error") +
  facet_wrap(~ latin, scale = "free_y")
```

## CPUE per length per haul
___

```{r}
common <- c("Clupea harengus", "Sprattus sprattus", "Gadus morhua",
            "Melanogrammus aeglefinus", "Merlangius merlangus", "Pollachius virens",
            "Trisopterus esmarkii", "Scomber scombrus", "Pleuronectes platessa")
cpue <- 
  # only valid hauls
  cpue_per_length_per_haul(hh = hh %>% filter(haulval == "V", !is.na(nsarea)),
                           hl = hl) %>% 
  filter(latin %in% common) %>% 
  # debugging, this should be fixed upstream
  filter(!is.na(length))
```

#### Comparison Datras service product

```{r, message = FALSE, warning = FALSE}
cpue2 <-
  cpue %>%
  id_separate(remove = FALSE) %>%
  filter(year %in% 2005:2016) %>% 
  select(id, latin, length, cpue.tidy = n) %>% 
  mutate(length = as.integer(length * 10))
fi <- 
  "data-raw/CPUE per length per haul per hour_2018-06-30 15_34_50.csv"
cpue.datras.service.product <-
  read_csv(fi, na = "-9") %>% 
  rename_all(tolower)
cpue.datras.service.product <-
  cpue.datras.service.product %>%
  filter(year %in% 2005:2016) %>% 
  id_unite() %>% 
  select(id, latin = species, length = lngtclass, cpue.dsp = cpue_number_per_hour) %>% 
  as_tibble()

diff <-
  cpue2 %>% 
  full_join(cpue.datras.service.product) %>%
  mutate(diff = !near(cpue.dsp, cpue.tidy, tol = 1e-4),
         txt = ifelse(!diff, "Same",
                      ifelse(diff, "Difference", "One of the value is NA")))
diff %>% 
  group_by(txt) %>% 
  summarise(n.records = n()) %>% 
  knitr::kable()
```

We have 515584 records that are OK and then 953 records that are different. We have 1642 values that are "NA" meaning that either of the cpue or both were missing.

```{r}
x <- diff %>% filter(is.na(txt))
table(!is.na(x$cpue.tidy), !is.na(x$cpue.dsp))
```

So have 1374 observations that are in "tidy" but not in "dsp" and 268 observations in "dsp" that are not in "tidy"

**Lets print the 268 observations that are not in "tidy"**:

```{r}
diff %>% 
  filter(is.na(txt), !is.na(cpue.dsp)) %>% 
  arrange(latin) %>% 
  knitr::kable()
```

What is the story with the single haddock?:
```{r}
raw <- read_rds("data-raw/datras/ns-ibts_raw.rds")
hl_raw <-
  raw$hl %>% 
  rename_all(tolower)
hl_raw %>% 
  as_tibble() %>% 
  id_unite() %>% 
  filter(id == "2009_1_THA2_GOV_26",
         valid_aphia == 126437) %>% glimpse()
```

So, it does not appear in the hl_raw dataframe. And because these are all of `lngtcode == "0"` these fish are what is termed as 0.5 cm length class, reporting units of millimeter. For the time being this does not quite make sense in my head.

What do we have in tidy for the haul and species in question?:
```{r}
cpue2 %>% 
  filter(id == "2009_1_THA2_GOV_26",
         latin == "Melanogrammus aeglefinus") %>% glimpse()
```

So no 20 cm fish

And in the dsp:
```{r}
cpue.datras.service.product %>% 
  filter(id == "2009_1_THA2_GOV_26",
         latin == "Melanogrammus aeglefinus") %>% glimpse()
```

We have only 20, 30 and 40 cm fish. So it seems like in the dsp the length class are treated as millimeters that are then converted to centimeters by rounding.

### Some simple summaries

```{r, message = FALSE, warning = FALSE}
d <-
  cpue %>% 
  mutate(b = n * 1e-5 * length^3) %>% 
  group_by(id, latin) %>% 
  summarise(n = sum(n),
            b = sum(b)) %>%
  separate(id, c("year", "quarter", "ship", "gear", "haulno"), convert = TRUE) %>%
  mutate(year = year + quarter/4)

d %>% 
  ggplot(aes(year, n, colour = factor(quarter))) +
  stat_summary(fun.data = "mean_cl_boot", size = 0.1) +
  expand_limits(y = 0) +
  facet_wrap(~ latin, scale = "free_y") +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = NULL,
       colour = "Quarter",
       title = "Annual abundance",
       subtitle = "Bootstrap mean and confidence interval of individual standardized tow duration [60 minutes]")
d %>% 
  ggplot(aes(year, b, colour = factor(quarter))) +
  stat_summary(fun.data = "mean_cl_boot", size = 0.1) +
  expand_limits(y = 0) +
  facet_wrap(~ latin, scale = "free_y") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(name = NULL, breaks = seq(1960, 2020, by = 10)) +
  scale_y_continuous(name = NULL, labels = NULL) +
  labs(colour = "Quarter",
       title = "Annual biomass",
       subtitle = "Bootstrap mean and confidence interval of individual standardized tow duration [60 minutes]")
```


```{r, message = FALSE, warning = FALSE}
d <-
  cpue %>%
  mutate(b = n * 0.01 * length^3) %>% 
  id_separate() %>% 
  group_by(year, quarter, latin, length) %>% 
  summarise(n = sum(n),
            b = sum(b)) %>% 
  ungroup()
d %>%
  filter(latin == "Melanogrammus aeglefinus",
         year >= 1998,
         quarter == 1,
         between(length, 9, 45)) %>% 
  ggplot(aes(length,  as.factor(year), group = year, height = n/1e5)) +
  geom_ridgeline(stat = "identity", scale = 5, alpha = 0.5)
d %>%
  filter(latin == "Gadus morhua",
         year >= 2000,
         quarter == 1,
         between(length, 5, 100)) %>% 
  ggplot(aes(length,  as.factor(year), height = n/1e3)) +
  geom_ridgeline(stat = "identity", scale = 3, alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 120, by = 10))
d %>%
  filter(latin == "Pleuronectes platessa",
         year >= 2000,
         quarter == 1) %>% 
  ggplot(aes(length, as.factor(year), height = n/1e3)) +
  geom_ridgeline(stat = "identity", alpha = 0.5)
```


## Field names
___

* **StNo**: National coding system, not defined by ICES
* **HaulNo**: Sequential numbering of hauls during cruise. In CA-records: HaulNo = -9 for Area-based ALK HaulNo <> -9 and > 0 for Haul-based ALK
* **DataType**:
    - -9: Invalid haul
    - C: Data calculated as CPUE (number per hour)
    - R: Data by haul
    - S: Subsample data
* **SpecCodeType**:
    - N: NODC code
    - T: TSN code
    - W: WoRMS ApiaID code
* **LngtCode**:
    - .:	1 mm length class, reporting units: mm
    - 0:	0.5 cm length class, reporting units: mm
    - 1:	1 cm length class, reporting units: cm
* **CatIdentifier**: Category for subsampling for species, length, weight, and sex. If **DataType** is C, **CatIdentifier** is always 1. For further description and examples of use look up the manual.
* **SubFactor**: Sub-sampling factor. If 1/6 of the catch was measured, report 6. If **DataType** is:
    - C: it should be reported as 1. 
    - S: it is always >1.
    - R: the SubFactor is = 1 or >1 for different species depending on whether they were subsampled.
